Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-06-26 review_article.Rmd
mapped out fir section

---
title: "Cartogram Mapping and Application to Cancer Data."
output: html_document
fig_caption: yes
number_sections: false
fontsize: 11pt
bibliography: reviewbib.bib
biblio-style: vancouver
link-citations: true
---

@@

### Choropleth Map (Steff)

Choropleth mapping methods effectively explore a relationship across a geographical space. 
Thematic maps involve drawing administrative or other boundaries and filling the regions with colours, symbols or patterns to present levels, or rates, of the information communicated. @CIBMUK believes these 'representations are highly intuitive' as the familiar shapes help those looking at the maps to visually infer spatial relationships. Using the topography as a base map is common for thematic maps of phenomena [@CD].
The visualisations are true maps, used for visual inspection and can help to direct hypotheses about the spatial patterns across the physical space. 
These true maps rely on visual perception, a fallible metric as it can cause viewers to 'weight the shades from the largest areas more heavily.' The inferences derived from visual inspection, and the hypotheses developed can suffer from 'bias'. The use of thematic disease mapping was motivated by a 'desire to evaluate geographical patterns' @DMAHP. This technique has been seen as a 'trigger for questions' which drives research of spatial distributions of diseases. @CPISACA recommends Choropleths, or isopleths for 'measures that are continuous over space, such as elevation or temperature.' @MACM also recognise the intuition that can be derived from choropleths with the critique that they must be 'validated by rigorous statistical analyses.'

In epidemiology, choropleths are often used as a mapping tool in the study of the spatial distribution. 
Almost 100 years of cancer maps have explored the US and the UK with increased effectiveness in presentation of unbiased rates. 
@HEDP considered maps a research tool, dots maps use spots, choropleths and thematic maps shade areas and isopleths use isolines. These  techniques were commonly used before the advent of computer-assisted cartography.
The data collection methods of cancer mortality rates across regions, and administrative control within regions lends itself choropleth visualisation. Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the proportion of old people. Howe [-@HEDP] describes Stocks development of the standardised mortality ratios through the 1930s. The choropleth maps presented levels of cancer via hatchings on a black and white scale.

@DMAHP cites Cruickshank's (1947) discussion of using visuals as 'formal statistical assessment of the spatial pattern' as a major advancement.
@CD considers cartographic data analysis in practice. Presenting a clear guide for preparing cartograms for use with both qualitative and quantitative information.
The authors recognise that the creation of cartograms was largely the work of professional cartographers until the innovation of geographical information systems that welcomed 
map users as map creators. These systems are utilised depending on 'the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994)'.
@SE reports the increase in disease mapping development due to the availability of geographic information system softwares.

@@

@@

Cartograms are a thematic map, presenting statistical and geographical information like choropleths. @ACTUC suggests the difference is the intentional and desirable augmentation of the size, shape or distance of geographical areas. World projections reflect the frequent distortions seen from altering perspectives. Choropleth maps will always be distorted if they cover enough of the globe, as will photographs of the globe from space. Cartogram creation requires choosing a favourable distortion of the geography for presenting the set of spatial information.
This concept was originally implemented to allow a map base that represented the proportion of the population within each region. @ACTUC considers this design a 'more socially just form of mapping' by giving all members of the population 'equitable representation'. This is known as an iso-demographic map, and the design inherently lends itself to epidemiolgy. 
@HEDP argues that 'cancer occurs in people, not in geographical areas' and the map base should reflect this to avoid allocating 'undue prominence' to rural areas. @DMAHP reduce the visual impact of large areas with small populations, the distortion of spatial areas is proportional to the denominator variable.

@@

@@

@@

@@

>>>>

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
output: html_document
fig_caption: yes
number_sections: false
fontsize: 11pt
bibliography: reviewbib.bib
biblio-style: vancouver
link-citations: true
---

@@

### Choropleth map (Steff)

<!-- what is a choropleth map -->
A choropleth map is used to show the spatial variability of measurements, with usage dating back to the 1800s. They are constructed by drawing administrative boundaries, and filling the polygon with colour to represent values of a measured variable. Early versions used symbols or patterns instead of colour. @CD provides a good overview (XXX check this is true), @CPISACA discusses the use of choropleth maps for visualising cancer data, and @DMAHP gives an overview of choropleth maps for displaying disease data. Figure 1 shows a choropleth map on XXX. What can be seen.  A choropleth map is a true map, constructed for visual inspection of spatial patterns across a familiar geographic form, that might show trends in disease occurrence, or even localised outbreaks.

< Figure 1 here: choropleth map >

<!-- psychology -->
The familiar shapes help viewers to visually infer spatial relationships, and hence intuitive to read [@CIBMUK]. However, the different sizes of administrative areas can lead viewers to place more attention on the shades of the largest areas. The inferences derived, and the hypotheses developed can suffer from a large area "bias". <!--This technique has been seen as a 'trigger for questions' which drives research of spatial distributions of diseases.  @MACM also recognise the intuition that can be derived from choropleths with the critique that they must be 'validated by rigorous statistical analyses.' -->

<!-- Data shown -->
In epidemiology, choropleths are often used as a mapping tool in the study of the spatial distribution. 
Almost 100 years of cancer maps have explored the US and the UK with increased effectiveness in presentation of unbiased rates. 
@HEDP considered maps a research tool, dots maps use spots, choropleths and thematic maps shade areas and isopleths use isolines. These  techniques were commonly used before the advent of computer-assisted cartography.
The data collection methods of cancer mortality rates across regions, and administrative control within regions lends itself choropleth visualisation. Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the proportion of old people. Howe [-@HEDP] describes Stocks development of the standardised mortality ratios through the 1930s. The choropleth maps presented levels of cancer via hatchings on a black and white scale.

<!-- Usage and availability -->
@DMAHP cites Cruickshank's (1947) discussion of using visuals as 'formal statistical assessment of the spatial pattern' as a major advancement.
@CD considers cartographic data analysis in practice. Presenting a clear guide for preparing cartograms for use with both qualitative and quantitative information.
The authors recognise that the creation of cartograms was largely the work of professional cartographers until the innovation of geographical information systems that welcomed 
map users as map creators. These systems are utilised depending on 'the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994)'.
@SE reports the increase in disease mapping development due to the availability of geographic information system softwares.

@@

<!-- Published atlases -->

@@

<!-- Going from choropleth to cartogram -->
Cartograms are a thematic map, presenting statistical and geographical information like choropleths. @ACTUC suggests the difference is the intentional and desirable augmentation of the size, shape or distance of geographical areas. World projections reflect the frequent distortions seen from altering perspectives. Choropleth maps will always be distorted if they cover enough of the globe, as will photographs of the globe from space. Cartogram creation requires choosing a favourable distortion of the geography for presenting the set of spatial information.
This concept was originally implemented to allow a map base that represented the proportion of the population within each region. @ACTUC considers this design a 'more socially just form of mapping' by giving all members of the population 'equitable representation'. This is known as an iso-demographic map, and the design inherently lends itself to epidemiolgy. 
@HEDP argues that 'cancer occurs in people, not in geographical areas' and the map base should reflect this to avoid allocating 'undue prominence' to rural areas. @DMAHP reduce the visual impact of large areas with small populations, the distortion of spatial areas is proportional to the denominator variable.

<!-- Common variables used to create cartograms, eg population, mortality -->

<!-- Overview of varieties -->

@@

< Figure 2 here - basic cartogram - easy to see >

@@

<!-- What is a contiguous cartogram -->

@@

< Figure 3 here>

@@

Dot plot: one dot for each region, coloured, and placed at centroid.
